---
title: "Appendix: Route selection analyses"
author:
- Hester Brønnvik  
- Department of Migration, Max Planck Institute of Animal Behavior, Radolfzell, Germany  
- Department of Biology, University of Konstanz, Konstanz, Germany
output: 
  pdf_document:
    highlight: haddock
  
always_allow_html: true
fontsize: 11
editor_options:
  markdown:
    wrap: 72
---

## Performing analyses using Step Selection Functions

Step-selection functions model animal movement as a series of discrete steps between consecutive locations. Each observed step is matched with alternative steps, which allows comparisons of the observed locations to alternative, available ones. Here, we generate these alternative steps and available locations and compare the available airflow. 

Step-selections functions define availability with three criteria: the previous location, the time between consecutive locations, and the characteristics of an individual's movement. First, we have to clean the data and process them into tracks. Then, we have to re-sample them so that there are consistent time intervals between locations. Finally, we have to extract the step lengths and turn angles for each individual and generate distributions of what is possible. 

Once we have generated the available locations, we can label them with environmental information using Movebank's Env-DATA service. The last step is to ask how these environmental factors predict space use, which we do by fitting conditional logistic regressions. 

### 1. Prepare the environment

The first step is to make sure that we have all of the packages and functions that we need. Here, we will use the "amt" package to make and process our tracks and the "survival" package to model our data. We need several additional packages for formatting, mapping, and ease.

----
```{r libraries, message=FALSE, warning=FALSE, results='hide'}
#set a working directory and load required packages
setwd("C:/Users/Tess Bronnvik/Desktop/Br-nnvik_honey_buzzard_ssf")
mypath <- paste0(getwd(),"/")
ssf_packs <- c("lubridate", "amt", "purrr", "move", "mapview", "ggpubr", "survival",
               "sjPlot", "MASS", "modelr", "mosaic", "tidyverse")
lapply(ssf_packs, require, character.only = TRUE)

```

```{r functions, message=FALSE, warning=FALSE, results='hide'}
#import required functions
NCEP.loxodrome.na <- function (lat1, lat2, lon1, lon2) {
  deg2rad <- pi/180
  acot <- function(x) {
    return(atan(1/x))
  }
  lat1 <- deg2rad * lat1
  lat2 <- deg2rad * lat2
  lon1 <- deg2rad * lon1
  lon2 <- deg2rad * lon2
  deltaLon <- lon2 - lon1
  pi4 <- pi/4
  Sig1 <- log(tan(pi4 + lat1/2))
  Sig2 <- log(tan(pi4 + lat2/2))
  deltaSig <- Sig2 - Sig1
  if (deltaLon == 0 && deltaSig > 0) {
    head <- 0
  }
  else if (deltaLon == 0 && deltaSig < 0) {
    head <- 180
  }
  else if (deltaSig == 0 && deltaLon > 0) {
    head <- 90
  }
  else if (deltaSig == 0 && deltaLon < 0) {
    head <- 270
  }
  else if (deltaSig < 0 && deltaLon < 0) {
    head <- acot(deltaSig/deltaLon) * 180/pi + 180
  }
  else if (deltaSig < 0 && deltaLon > 0) {
    head <- acot(deltaSig/deltaLon) * 180/pi + 180
  }
  else if (deltaSig > 0 && deltaLon > 0) {
    head <- acot(deltaSig/deltaLon) * 180/pi
  }
  else if (deltaSig > 0 && deltaLon < 0) {
    head <- acot(deltaSig/deltaLon) * 180/pi + 360
  }
  else {
    head <-NA}
  return(head)
}
source(paste0(mypath, "wind_support_Kami.R"))

```
----

Next, we set some criteria for building the tracks. The step number determines how many alternative steps are generated for each observed one. The tolerance determines how many minutes around a fix are allowed in order to consider it a next step (eg. a step is the distance traveled in 2 hours +/- 15 minutes). Finally, the CRS is the coordinate reference system, defining the map projection for the mk_track function. Here we have to inform R that our data are measured in degrees rather than in meters. 

----
```{r criteria}
# set criteria for tracks
stepNumber <- 100 # random steps
toleranceLength <- 15 # tolerance in minutes
wgs <- CRS("+proj=longlat +datum=WGS84 +no_defs") # map projection

```
----

### 2. Prepare data for Movebank's environmental data annotation

In order to get the data in the right format, we select only the reads from autumn migrations and clean them. We need to remove any duplicated reads and any times when the birds were not moving. We also need to map the data to see whether there are any incomplete records, and to remove those records if we find any. Then we need to separate the data so that individuals that were sampled at different rates are analyzed separately.

----
```{r data}
# retrieve the full data set
full_data <- read.csv(paste0(mypath,"original_data/full_buzz_set.csv"),stringsAsFactors = F, header = T)

# reformat the time stamps
full_data$timestamp <- as.POSIXct(strptime(full_data$dt, format = "%Y-%m-%d %H:%M:%S"),
                                  tz = "UTC")

# select the autumn migrations
all_autumns <- full_data[grep("autumn", full_data$phase),]

# find and remove duplicate observations
doubles <- all_autumns %>% dplyr::select(long, lat, name, timestamp) %>%
  duplicated
sum(doubles) # 59 duplicates
all_autumns <- all_autumns[doubles != TRUE,]

# get the 204 reads when the birds did not move and remove them from the data
resting_reads <- read.csv(paste0(mypath,"resting.csv"), stringsAsFactors = F)
all_autumns <- all_autumns %>% mutate(id_ts = paste(name,timestamp, sep="_")) %>% 
  filter(!id_ts %in% resting_reads$id_ts)

# order all the data by timestamp
all_autumns <- all_autumns %>%
  arrange(timestamp)

# remove birds of unknown age with fewer than 4 tracked autumn migrations
all_autumns <- all_autumns %>% filter(all_autumns$name != "Aarne" &
                         all_autumns$name != "Jari" &
                         all_autumns$name != "Johannes" &
                         all_autumns$name != "Kari")

# create a unique identifier for each migratory journey
all_autumns$id_year <- paste(all_autumns$name, all_autumns$yr, sep="_")

# make simple plots of lat/long to check for outliers
ggplot(all_autumns, aes(x=long, y=lat, color=as.factor(name))) + geom_point() +
  theme(legend.position = "none")
ggplot(all_autumns, aes(x=long, y=lat)) + geom_point()+ facet_wrap(~name, scales="free")

# make a map of individual tracks
data_sp <- all_autumns # store the data, then create a spatial object for plotting
coordinates(data_sp) <- ~ long + lat # set coordinates
proj4string(data_sp) <- wgs # set projection
mapView(data_sp, zcol = "id_year", burst = F, cex = 3, color = rainbow) # plot on a map

# some individuals do not survive their first trip and have to be removed from the data
all_autumns <- all_autumns %>% filter(name != "Emma" &
                                      name != "Lisa" &
                                      name != "Miikka" &
                                      name != "Per" &
                                      name != "Sven" &
                            # others died en route later in life and must also be removed
                                      id_year != "Annika_2015" & 
                                      id_year != "Jouko_2015" & 
                                      id_year != "Paivi_2017" &
                                      id_year != "Senta_2017")

# the data need to be re-sampled at different rates label them 
# with these rates in minutes so that each can be processed individually.
all_autumns$sample_rate <- NA

# the individuals with 1 hour sampling rates
all_autumns$sample_rate[which(all_autumns$name == "Anni" | 
                              all_autumns$name == "Viljo")] <- 60

# the individuals with 2 hour sampling rates
all_autumns$sample_rate[which(all_autumns$name == "Edit" | all_autumns$name == "Julia" |
                              all_autumns$name == "Matti"| all_autumns$name == "Ulla" |
                              all_autumns$name == "Aida" | all_autumns$name == "Ella" |
                              all_autumns$name == "Heidi" | all_autumns$name == "Kirsi" |
                              all_autumns$name == "Gilda" | 
                              all_autumns$name == "Valentin")] <- 120

# the individuals with 3 hour sampling rates
all_autumns$sample_rate[which(all_autumns$name == "Mohammed")] <- 180

# the individuals with 4 hour sampling rates
all_autumns$sample_rate[which(all_autumns$name == "Annika" | all_autumns$name == "Jaana"|
                              all_autumns$name == "Lars"| all_autumns$name == "Piff"| 
                              all_autumns$name == "Puff" | all_autumns$name == "Roosa"|
                              all_autumns$name == "Senta"| all_autumns$name == "Tor" |
                              all_autumns$name == "Tiina" |all_autumns$name == "Jouko" |
                              all_autumns$name == "Mikko" |all_autumns$name == "Paivi" |
                              all_autumns$name == "Hans" | all_autumns$name == "Venus" |
                              all_autumns$name == "Rudolf")] <- 240
```
----

Using the amt package, we can create track objects. These are then re-sampled to a given step length. Steps have to be of consistent times because the parameters of the models are scale dependent and will vary with $\Delta$t. By re-sampling, we burst the tracks. A burst is a track segment containing fixes separated by the given step length. We can then create randomized locations as part of these bursts.

----
```{r build-tracks, message=FALSE, warning=FALSE, results='hide'}
# build tracks at each rate

autumn_track <- data.frame()

for (i in sort(unique(all_autumns$sample_rate))) {
  # select data of the given sampling rate
  temp_rdata <- all_autumns[which(all_autumns$sample_rate == i),]
  for (j in unique(temp_rdata$id_year)) {
    # select data for the given track
    temp_idata <- temp_rdata[which(temp_rdata$id_year == j),]
  # make the track  
  trk <- mk_track(temp_idata,.x=long, .y=lat, .t=timestamp, id = name, crs = wgs)
  # resample to a consistent time between steps
  trk <- track_resample(trk, rate = minutes(i), tolerance = minutes(toleranceLength))
  # remove bursts with fewer than three re-locations so that turn angle can be calculated
  trk <- filter_min_n_burst(trk, 3)
  # burst steps
  burst <- steps_by_burst(trk, keep_cols = "start")
  # create random steps using fitted gamma and von Mises distributions and append
  rnd_stps <- burst %>% random_steps(n_control = stepNumber)
  # save
  autumn_track <- rbind(autumn_track,rnd_stps)
  # and signal
  print(paste0("Successfully created random steps for track ", j))
  }
}

# compare the step lengths and turn angles for observed and random steps
obs_sl <- autumn_track[autumn_track$case_ == TRUE,] %>% 
  ggplot(aes(sl_, fill = factor(id))) + ggtitle("Observed") + geom_density(alpha = 0.4) + theme_minimal()
obs_ta <- autumn_track[autumn_track$case_ == TRUE,] %>% 
  ggplot(aes(ta_, fill = factor(id))) + ggtitle("Observed") +
  geom_density(alpha = 0.4) + 
  labs(x = "Turn angle (radians)", y = "Density") + 
  theme_minimal()
rand_sl <- autumn_track[autumn_track$case_ == FALSE,] %>%  
  ggplot(aes(sl_, fill = factor(id))) + ggtitle("Alternative") + geom_density(alpha = 0.4) + theme_minimal()
rand_ta <- autumn_track[autumn_track$case_ == FALSE,] %>% 
  ggplot(aes(ta_, fill = factor(id))) + ggtitle("Alternative") +
  geom_density(alpha = 0.4) + 
  labs(x = "Turn angle (radians)", y = "Density") + 
  theme_minimal()
ggarrange(obs_sl, rand_sl, obs_ta, rand_ta, ncol=2, nrow=2, legend = "none")


```
----

In order to calculate crosswind and tailwind components of the north/south and east/west winds that we will get from Movebank annotation, we need to calculate the direction of movement (loxodrome angle of travel) between locations. This is best done before annotation to avoid truncation of smaller values in the locations during file export.

----
```{r headings}
# calculate directions of movement between locations for each step
autumn_track <- autumn_track %>% rowwise() %>% 
  mutate(heading = NCEP.loxodrome.na(y1_, y2_, x1_, x2_)) %>% 
  ungroup()
```
----

Now we only need to conform to column name and file size specifications, then export the prepared data and send it to the Env-DATA service.

----
```{r export}
# select the end point for each step
names(autumn_track)[c(2,4)] <-c("location-long", "location-lat") 
# select the start time for each step
autumn_track$timestamp<-autumn_track$t1_ 
autumn_track$timestamp <- paste0(autumn_track$timestamp,".000" )

# split the data to conform to Movebank's file size limits
half1 <- autumn_track[1:(0.5*nrow(autumn_track)),]
half2 <- autumn_track[(0.5*nrow(autumn_track)+1):nrow(autumn_track),]

# export each file for Movebank annotation
write.csv(half1, paste0("HB1_",Sys.Date(),".csv"), row.names = FALSE)
write.csv(half2, paste0("HB2_",Sys.Date(),".csv"), row.names = FALSE)
```
----

### 3. Prepare data for modeling

Step-selection analyses use conditional logistic regressions to predict whether a location is used or unused based on environmental conditions. Once we have our files returned from Movebank's Env-DATA service, we can process the annotated data for modeling. Using the directions we calculated above, we can derive the crosswind and tailwind components from the wind data. Our third variable of interest is the vertical velocity of pressure, which proxies uplift and subsidence. Because pressure is lower with increasing altitude, negative vertical velocity values indicate uplift. 

----
```{r import, results='hide', message=FALSE, warning=FALSE}
# read in the annotated data
half1 <- read.csv(paste0(mypath,"HB1_4433398109146553161.csv"),
                  stringsAsFactors = F, header = T)
half2 <- read.csv(paste0(mypath,"HB2_4573417134914709132.csv"),
                  stringsAsFactors = F, header = T)
# and bind them into a single frame
annotated_data <- rbind(half1,half2) %>%
  # rename the columns to be more useful
  rename(lon1 = x1_,
         lat1 = y1_,
         lon2 = location.long,
         lat2 = location.lat,
         u_wind = ECMWF.ERA5.PL.U.Wind,
         v_wind = ECMWF.ERA5.PL.V.wind,
         vertical_pressure = ECMWF.ERA5.PL.Pressure.Vertical.Velocity) %>% 
  # then reformat the time and logical, and calculate wind support and crosswind
  mutate(timestamp = as.POSIXct(strptime(timestamp, format = "%Y-%m-%d %H:%M:%S"),
                                tz = "UTC"),
         year = format(as.POSIXct(timestamp,format="%Y-%m-%d %H:%M:%S"),"%Y"),
         id_year = paste(id,year,sep="_"),
         case_ = as.numeric(case_),
         tail = wind_support(u_wind, v_wind, heading),
         cross = cross_wind(u_wind, v_wind, heading))

```
----

As a last step we standardize our predictors across the whole data set. This means that all of our data have the same mean and variance, and therefore allows us to compare the coefficients of different models. We use "scale" to divide each value by its column's standard deviation and subtract the mean. We also take the absolute value of the crosswind variable because sign signifies compass direction, but we are only interested in the strength of the wind. 

----
```{r standardize}
#standardize the predictors
annotated_data <- annotated_data %>% 
  mutate(scaled_cross = abs(as.numeric(scale(cross, center = T, scale = T))), 
         scaled_tail = as.numeric(scale(tail, center = T, scale = T)),
         scaled_vertical = as.numeric(scale(vertical_pressure, center = T, scale = T)),)

```
----

### 4. Build the model

We need to model each individual in each year separately, which means fitting as many models as there are tracks. We can do this quickly using nested lists if we first write our model as a function and then give it to the "purrr" package's "map" function.

----
```{r modeling}
# write the model function
modelTCV <- function(df) {
  clogit(case_ ~ scaled_tail + scaled_cross + scaled_vertical + ta_ + sl_ + strata(step_id_), data = df)
}
#check how many true steps survived filtering
print(annotated_data %>% group_by(id_year) %>% tally(case_ == 1), n= 70)

# fit separate models for each journey, collect the coefficients, and calculate the AICs
SSF_results <- annotated_data %>% 
  group_by(id_year) %>% 
  nest() %>% 
  mutate(ssf_modelTCV = purrr::map(data, modelTCV),
         ssf_coefsTCV = purrr::map(ssf_modelTCV, coef),
         AIC_TCV = map_dbl(ssf_modelTCV, ~AIC(.)))

```
----

### 5. Extract the model coefficients for plotting

----
```{r extraction}
# flatten the coefficient column
ssf_coefs <- unnest(SSF_results, ssf_coefsTCV)

#the coefficients for each variable are now in a single column, group them by their identities
tail_coefs <- do.call(rbind, lapply(unique(ssf_coefs$id_year),
                      function(x){head(ssf_coefs[ssf_coefs$id_year == x,], 3)[1,]}))
cross_coefs <- do.call(rbind, lapply(unique(ssf_coefs$id_year),
                        function(x){head(ssf_coefs[ssf_coefs$id_year == x,], 3)[2,]}))
lift_coefs <- do.call(rbind, lapply(unique(ssf_coefs$id_year),
                        function(x){tail(ssf_coefs[ssf_coefs$id_year == x,], 3)[3,]}))

#create a common label for the variables
tail_coefs$var <- "Wind support"
cross_coefs$var <- "Crosswind"
lift_coefs$var <- "Vertical velocity"


#bind them into a single object again with their identities appended
plot_data <- rbind(tail_coefs, cross_coefs,lift_coefs)
```
----

To plot data in groups, we add the life stage and migration number to each individual.

----
```{r id_year}
#separate the id_year column so that stage is identified by individual
plot_data <- separate(plot_data, col = "id_year", c("id", "year"), remove = F)
plot_data$stage <- NA # create an empty column to store values
plot_data$stage[which(plot_data$id == "Annika" |
                         plot_data$id == "Jouko" |
                         plot_data$id == "Mikko" |
                         plot_data$id == "Paivi" |
                         plot_data$id == "Tiina" )] <- "adult" # label individuals of unknown age
plot_data$stage[which(is.na(plot_data$stage))] <- "juvenile" # label individuals tagged as fledglings

#add the number of each migration
plot_data$Migration <- NA 
plot_data$Migration[which(plot_data$id_year == "Jaana_2013" | # add year 2 info
                             plot_data$id_year == "Lars_2013" | 
                             plot_data$id_year == "Mohammed_2018" | 
                             plot_data$id_year == "Senta_2015" | 
                             plot_data$id_year == "Valentin_2015" |
                             plot_data$id_year == "Annika_2012" | 
                             plot_data$id_year == "Jouko_2012" |
                             plot_data$id_year == "Paivi_2014" |
                             plot_data$id_year == "Mikko_2012" |
                             plot_data$id_year == "Tiina_2012")] <- "2"

plot_data$Migration[which(plot_data$id_year == "Jaana_2014" | # add year 3 info
                             plot_data$id_year == "Lars_2014" | 
                             plot_data$id_year == "Senta_2016" |
                             plot_data$id_year == "Annika_2013" |
                             plot_data$id_year == "Jouko_2013" |
                             plot_data$id_year == "Mikko_2013" |
                             plot_data$id_year == "Paivi_2015" |
                             plot_data$id_year == "Mohammed_2019" |
                             plot_data$id_year == "Tiina_2013")] <- "3"

plot_data$Migration[which(plot_data$id_year == "Lars_2015" | # add year 4 info
                             plot_data$id_year == "Mohammed_2020")] <- "4"

plot_data$Migration[which(plot_data$id_year == "Annika_2014" | # add year 5 info
                             plot_data$id_year == "Jouko_2014" |
                             plot_data$id_year == "Mikko_2014" | 
                             plot_data$id_year == "Mikko_2015" |
                             plot_data$id_year == "Paivi_2016" |
                             plot_data$id_year == "Tiina_2014" |
                             plot_data$id_year == "Tiina_2015")] <- "5 +" 
plot_data$Migration[which(is.na(plot_data$Migration))] <- "1" # everything else is year 1

plot_data <- plot_data[which(plot_data$stage == "juvenile" | plot_data$stage == "adult" & plot_data$Migration == "5 +"),]

```
----

### 6. Plot the coefficients

First, select the data from the stage and variable of interest. Then, select the data from individuals that survived in multiple years and remove the multiple year individuals from the set containing the rest. Plot a box plot with dashed lines for all the individuals, and then layer on a box plot with solid lines and without whiskers. Add points for birds surviving only one year. Layer on points for birds surviving in multiple years and connect those points with lines. Finally plot all of this in panels with separate axes.

----
```{r boxplots, fig.height = 5, fig.width = 7}
# select colors that are distinct
id_colors <- c("#542344","#1478A3","#00A354","#F34213","#FFBF00")

# select data from one variable and life stage
tail_plot_data <- plot_data %>% filter(var == "Wind support")
# create a new variable that contains the individuals surviving multiple migrations
tail_plot_highlight <- tail_plot_data %>% filter(id == "Jaana" | id == "Senta" | 
                                                 id == "Valentin" | id == "Mohammed" | 
                                                 id == "Lars")
# create a new variable with only individuals that died before a second migration 
tail_plot_data_points <- tail_plot_data %>% filter(id != "Jaana" & id != "Senta" & 
                                                   id != "Valentin" & id != "Mohammed" & 
                                                   id != "Lars")
# make the plot
tail_plot <- ggplot(tail_plot_data, aes(x=Migration, y=ssf_coefsTCV)) + 
  # a box plot with all the data
  geom_boxplot(outlier.shape = NA, color = "grey30", linetype="dashed")+ 
  # a second box plot without whiskers
  stat_boxplot(aes(ymin = ..lower.., ymax = ..upper..), outlier.shape = NA, fill = "grey80",
               lwd = 0.75, fatten = 1.2) +
  # the points for individuals that died
  geom_jitter(data = tail_plot_data_points, aes(), size=2, width=0.3, color = "grey30") +
  # the points for individuals with multiple migrations
  geom_point(data = tail_plot_highlight, aes(x=Migration, y=ssf_coefsTCV, fill = id), 
             shape = 21, size = 3) +
  # the lines connecting the repeated journeys
  geom_line(data = tail_plot_highlight, aes(group = id, color =id), size = 1.2) +
  # color the points
  scale_fill_manual(values = id_colors) + 
  # color the lines
  scale_color_manual(values = id_colors) + 
  # label the plot and axes
  labs(title = "Wind support", x="Migration", y = "Coefficients" ) +
  # make a line at y = 0 for clarity
  geom_hline(yintercept = 0) +
  # remove background grid, make axis labels black, and increase text size
  theme_classic() +
    theme(legend.position = "none", text = element_text(size=15), 
        axis.text.x = element_text(colour = "black"),
        axis.text.y = element_text(colour = "black"))

# repeat the above for plots of the other predictor variables
cross_plot_data <- plot_data %>% filter(var == "Crosswind")
cross_plot_highlight <- cross_plot_data %>% filter(id == "Jaana" | id == "Senta" |
                                                   id == "Valentin" | id == "Mohammed" | 
                                                   id == "Lars")
cross_plot_data_points <- cross_plot_data %>% filter(id != "Jaana" & id != "Senta" & 
                                                     id != "Valentin" & id != "Mohammed" &
                                                     id != "Lars")

cross_plot <- ggplot(cross_plot_data, aes(x=Migration, y=ssf_coefsTCV)) + 
  geom_boxplot(outlier.shape = NA, color = "grey30", linetype="dashed")+
  stat_boxplot(aes(ymin = ..lower.., ymax = ..upper..), outlier.shape = NA, fill = "grey80",
               lwd = 0.75, fatten = 1.2) +
  geom_jitter(data = cross_plot_data_points, aes(), size=2, width=0.3, color = "grey30") +
  geom_point(data = cross_plot_highlight, aes(x=Migration, y=ssf_coefsTCV, fill = id), 
             shape = 21, size = 3)  + 
  geom_line(data = cross_plot_highlight, aes(group = id, color =id), size = 1.2) + 
  scale_fill_manual(values = id_colors) + 
  scale_color_manual(values = id_colors) + 
  labs(title = "Crosswind", x="Migration", y = NULL ) +
  geom_hline(yintercept = 0) +
  theme_classic() +
  theme(legend.position = "none", text = element_text(size=15), 
        axis.text.x = element_text(colour = "black"),
        axis.text.y = element_text(colour = "black"))

lift_plot_data <- plot_data %>% filter(var == "Vertical velocity")
lift_plot_highlight <- lift_plot_data %>% filter(id == "Jaana" | id == "Senta" | 
                                                 id == "Valentin" | id == "Mohammed" | 
                                                 id == "Lars")
lift_plot_data_points <- lift_plot_data %>% filter(id != "Jaana" & id != "Senta" & 
                                                   id != "Valentin" & id != "Mohammed" & 
                                                   id != "Lars")

lift_plot <- ggplot(lift_plot_data, aes(x=Migration, y=ssf_coefsTCV)) + 
  geom_boxplot(outlier.shape = NA, color = "grey30", linetype="dashed")+
  stat_boxplot(aes(ymin = ..lower.., ymax = ..upper..), outlier.shape = NA, fill = "grey80",
               lwd = 0.75, fatten = 1.2) +
  geom_jitter(data = lift_plot_data_points, aes(), size=2, width=0.3, color = "grey30") +
  geom_point(data = lift_plot_highlight, aes(x=Migration, y=ssf_coefsTCV, fill = id), 
             shape = 21, size = 3)  + 
  geom_line(data = lift_plot_highlight, aes(group = id, color =id), size = 1.2) + 
  scale_fill_manual(values = id_colors) + 
  scale_color_manual(values = id_colors) + 
  labs(title = "Vertical velocity", x="Migration", y = NULL ) +
  geom_hline(yintercept = 0) +
  theme_classic() +
  theme(legend.position = "none", text = element_text(size=15), 
        axis.text.x = element_text(colour = "black"),
        axis.text.y = element_text(colour = "black"))

ggarrange(tail_plot, cross_plot, lift_plot, ncol=3, nrow=1, legend = "none")
```

The importance of wind support, crosswind, and uplift to route selection over migrations. Negative values of vertical velocity correspond to rising air. We fitted separate conditional logistic regressions for each individual in each migration and extracted the coefficients for each predictor. Coefficients of individuals surviving in multiple years are connected by lines and highlighted in color.



## Permutations and significance testing 

Because we have only one data set, we cannot compare our result to the results from similar systems: we have no other data sets of the same size, with the same sampling effort, and from populations with the same means and variances of the parameters of interest. However, we can generate alternative data sets that meet these requirements and that also meet a null expectation. Then, we can compare our observed effects to the effects in these alternative data. To do this, we repeatedly randomize our response variable with respect to our predictor variables, model these randomized data, and store the results. The outcome is a distribution of results that would occur if there were no relationship between the dependent and independent variables in our model. 

### 1. Prepare the data for modeling

Here, we use the "mosaic" package to perform the permutations because it allows us to maintain a stratified design in our models. 

---
```{r environment, message=FALSE, warning=FALSE, results='hide'}

# set the number of permutations
no.perm <- 101 # here it is equal to the number of used + available steps in each stratum

# label the data with life stage and migration information. This is a repetition of what was
# done above, but here we label the full data rather than the coefficients
annotated_data$stage <- NA
annotated_data$stage[which(annotated_data$id == "Annika" |
                             annotated_data$id == "Jouko" |
                             annotated_data$id == "Mikko" |
                             annotated_data$id == "Paivi" |
                             annotated_data$id == "Tiina" )] <- "adult"
annotated_data$stage[which(is.na(annotated_data$stage))] <- "juvenile"
                                
annotated_data$Migration <- NA # create an empty column to store values
annotated_data$Migration[which(annotated_data$id_year == "Jaana_2013" | # add year 2 info
                                 annotated_data$id_year == "Lars_2013" | 
                                 annotated_data$id_year == "Mohammed_2018" | 
                                 annotated_data$id_year == "Senta_2015" | 
                                 annotated_data$id_year == "Valentin_2015" |
                                 annotated_data$id_year == "Annika_2012" |
                                 annotated_data$id_year == "Jouko_2012" |
                                 annotated_data$id_year == "Paivi_2014" |
                                 annotated_data$id_year == "Mikko_2012" |
                                 annotated_data$id_year == "Tiina_2012")] <- "2"

annotated_data$Migration[which(annotated_data$id_year == "Jaana_2014" | # add year 3 info
                                 annotated_data$id_year == "Lars_2014" | 
                                 annotated_data$id_year == "Senta_2016" |
                                 annotated_data$id_year == "Annika_2013" |
                                 annotated_data$id_year == "Mikko_2013" |
                                 annotated_data$id_year == "Paivi_2015" |
                                 annotated_data$id_year == "Mohammed_2019" |
                                 annotated_data$id_year == "Tiina_2013")] <- "3"

annotated_data$Migration[which(annotated_data$id_year == "Lars_2015" | # add year 4 info
                                 annotated_data$id_year == "Mohammed_2020")] <- "4"

annotated_data$Migration[which(annotated_data$id_year == "Annika_2014" | # add year 5 info
                            annotated_data$id_year == "Jouko_2014" |
                            annotated_data$id_year == "Mikko_2014" | 
                            annotated_data$id_year == "Mikko_2015" |
                            annotated_data$id_year == "Paivi_2016" |
                            annotated_data$id_year == "Tiina_2014" |
                            annotated_data$id_year == "Tiina_2015")] <- "5 +" 
annotated_data$Migration[which(is.na(annotated_data$Migration))] <- "1" # everything else is year 1

# make a new unique vector for each step and migration to serve as strata during permutation
annotated_data$id_year_step <- paste(annotated_data$id_year, annotated_data$step_id_, sep="_")
```
---


### 2. Perform data stream permutation and test for significance

Once we have modeled all of our observed data, we can run our randomizations. Unfortunately, this becomes a repetitive process because each migration must be modeled separately, the significance of each variable in each migration must be calculated separately, and each distribution of coefficients must be plotted separately. We create an empty data frame so that we have a place to store our results. Then, we use the "mosaic" package to re-sample our data set. We hold our predictors in the same positions relative to the identifying data for each individual and to the other predictors. We shuffle the dependent variable within each group, which we define as a step containing one observation and 100 random locations. Next, we fit our model to the randomized data, extract the coefficients, and save them in our data frame. We repeat this process as many times as we want, which we defined as "no.perm" above. 

```{r permutations, message=FALSE, warning=FALSE, results=F}
#create an empty data frame to store the coefficients in
random_model <- data.frame()
# create a vector to identify migrations by
migrations <- unique(annotated_data$Migration)
# and a second, empty one that will act as a column in the data frame 
Migration <- c()
# create a loop to go through each migration
for (j in migrations) {
  # select the data from the given migration
  temp_data <- annotated_data[which(annotated_data$Migration == j),]
# then create a second loop to permute the data
  for (i in 1:no.perm) {
    # this is the workhorse, taking each migration and permuting within the strata
    rando <- mosaic::resample(temp_data, replace = F, groups = id_year_step, shuffled = "case_", 
                              fixed = c("id","year","scaled_cross","scaled_tail","scaled_vertical",
                                        "timestamp", "ta_", "sl_", "step_id_"))
    # model those permuted data sets
    temp_model <- modelTCV(rando)
    # extract the coefficients from the model
    temp_coefs <- summary(temp_model)$coefficients
    # store the coefficients in the data frame
    random_model <- rbind(random_model, temp_coefs)
    # add three more migration labels to identify the three coefficients
    Migration <- c(Migration,j,j,j)
    # and finally we let ourselves know we have done this
    print(paste0("Completed permutation ", i, "."))
  }
  print(paste0("Completed permutation of migration ", j, "."))
}
colnames(random_model)[1] <- "Estimate"
# add each migration label to the coefficients
random_model$Migration <- Migration
```
```{r model}
head(random_model)
```


---

Once we have aggregated all of the coefficients of the permutations for each migration, we compare the observed coefficients to the randomized ones to test for significance.

---
```{r significance}
## First extract the coefficients of the models for the observed data
# fit the same model as above, but now for each migration rather than each individual track
clogit_results <- annotated_data %>% 
  group_by(Migration) %>% 
  nest() %>% 
  mutate(model_col = purrr::map(data, modelTCV),
         coefsTCV = purrr::map(model_col, coef),
         AIC_TCV = map_dbl(model_col, ~AIC(.)),
         summary_stats = purrr::map(model_col, ~summary(.x))) %>% 
  unnest(cols = c(coefsTCV, Migration)) %>% 
  mutate(variable = c("tail","cross","vertical")) %>%
  arrange(variable)
clogit_results

# create a vector of the variables in clogit_results
variable <- unique(clogit_results$variable)
# and a data frame to store the significance values in
significances <- data.frame()

# then calculate the significance of the difference between each set of permuted coefficients 
# and the observed 
for (i in migrations) {
  # select the coefficients for migration i
  temp_observed <- clogit_results[which(clogit_results$Migration == i),]
  # and the coefficients from permuted data of migration i
  temp_random <- random_model[grepl(i, random_model$Migration),]
  for (j in variable) {
    # select the coefficient that is also variable j
    temp_obs <- temp_observed[which(temp_observed$variable == j),]
    # and the permuted coefficients of j
    temp_rand <- temp_random[grepl(j, row.names(temp_random)),]
    # count up the permuted coefficients with a more extreme value than the observed and 
    # divide by the number of coefficients to get the proportion
    significance <- sum(abs(temp_obs$coefsTCV) < abs(temp_rand$Estimate))/no.perm
    temp_sig <- c(i, j, significance)
    significances <- rbind(significances, temp_sig)
  }
}

colnames(significances) <- c("Migration", "Variable", "Significance")

significances
```
---

## 3. Plot the observed and randomized coefficients

Finally, we plot the distributions of coefficients derived from randomized data along with the observed coefficients.

---
```{r plots, message=FALSE, warning=FALSE, results='hide', out.extra='angle=90'}
no.bin <- 25 # set the number of bins for the histograms
first_tail <- ggplot(data = filter(random_model, 
                               grepl("tail", rownames(random_model)) & 
                               Migration == 1), aes(Estimate)) + 
  geom_histogram(fill="black", bins = no.bin) +
  geom_vline(aes(xintercept = filter(clogit_results, variable == "tail" & 
                               Migration == 1)$coefsTCV), color="red", size=1) +
  theme_classic()+
  theme(legend.position = "none", text = element_text(size=20), 
        axis.text.x = element_text(colour = "black"),
        axis.text.y = element_text(colour = "black"))
first_cross <- ggplot(data= filter(random_model, 
                                grepl("cross", rownames(random_model)) 
                                & Migration == 1), aes(Estimate)) + 
  geom_histogram(fill="black", bins = no.bin) +
  geom_vline(aes(xintercept = filter(clogit_results, variable == "cross" & 
                               Migration == 1)$coefsTCV), color="red", size=1) +
  theme_classic()+
  theme(legend.position = "none", text = element_text(size=20), 
        axis.text.x = element_text(colour = "black"),
        axis.text.y = element_text(colour = "black"))
first_lift <- ggplot(data= filter(random_model, 
                                grepl("vertical", rownames(random_model)) 
                                & Migration == 1), aes(Estimate)) + 
  geom_histogram(fill="black", bins = no.bin) +
  geom_vline(aes(xintercept = filter(clogit_results, variable == "vertical" & 
                               Migration == 1)$coefsTCV), color="red", size=1) +
  theme_classic()+
  theme(legend.position = "none", text = element_text(size=20), 
        axis.text.x = element_text(colour = "black"),
        axis.text.y = element_text(colour = "black"))

second_tail <- ggplot(data= filter(random_model, 
                               grepl("tail", rownames(random_model)) & 
                               Migration == 2), aes(Estimate)) + 
  geom_histogram(fill="black", bins = no.bin) +
  geom_vline(aes(xintercept = filter(clogit_results, variable == "tail" & 
                               Migration == 2)$coefsTCV), color="red", size=1) +
  theme_classic()+
  theme(legend.position = "none", text = element_text(size=20), 
        axis.text.x = element_text(colour = "black"),
        axis.text.y = element_text(colour = "black"))
second_cross <- ggplot(data= filter(random_model, 
                                grepl("cross", rownames(random_model)) 
                                & Migration == 2), aes(Estimate)) + 
  geom_histogram(fill="black", bins = no.bin) +
  geom_vline(aes(xintercept = filter(clogit_results, variable == "cross" & 
                               Migration == 2)$coefsTCV), color="red", size=1) +
  theme_classic()+
  theme(legend.position = "none", text = element_text(size=20), 
        axis.text.x = element_text(colour = "black"),
        axis.text.y = element_text(colour = "black"))
second_lift <- ggplot(data= filter(random_model, 
                                grepl("vertical", rownames(random_model)) 
                                & Migration == 2), aes(Estimate)) + 
  geom_histogram(fill="black", bins = no.bin) +
  geom_vline(aes(xintercept = filter(clogit_results, variable == "vertical" & 
                               Migration == 2)$coefsTCV), color="red", size=1) +
  theme_classic()+
  theme(legend.position = "none", text = element_text(size=20), 
        axis.text.x = element_text(colour = "black"),
        axis.text.y = element_text(colour = "black"))
  
third_tail <- ggplot(data= filter(random_model, 
                               grepl("tail", rownames(random_model)) & 
                               Migration == 3), aes(Estimate)) + 
  geom_histogram(fill="black", bins = no.bin) +
  geom_vline(aes(xintercept = filter(clogit_results, variable == "tail" & 
                               Migration == 3)$coefsTCV), color="red", size=1) +
  theme_classic()+
  theme(legend.position = "none", text = element_text(size=20), 
        axis.text.x = element_text(colour = "black"),
        axis.text.y = element_text(colour = "black"))
third_cross <- ggplot(data= filter(random_model, 
                                grepl("cross", rownames(random_model)) 
                                & Migration == 3), aes(Estimate)) + 
  geom_histogram(fill="black", bins = no.bin) +
  geom_vline(aes(xintercept = filter(clogit_results, variable == "cross" & 
                               Migration == 3)$coefsTCV), color="red", size=1) +
  theme_classic()+
  theme(legend.position = "none", text = element_text(size=20), 
        axis.text.x = element_text(colour = "black"),
        axis.text.y = element_text(colour = "black"))
third_lift <- ggplot(data= filter(random_model, 
                                grepl("vertical", rownames(random_model)) 
                                & Migration == 3), aes(Estimate)) + 
  geom_histogram(fill="black", bins = no.bin) +
  geom_vline(aes(xintercept = filter(clogit_results, variable == "vertical" & 
                               Migration == 3)$coefsTCV), color="red", size=1) +
  theme_classic()+
  theme(legend.position = "none", text = element_text(size=20), 
        axis.text.x = element_text(colour = "black"),
        axis.text.y = element_text(colour = "black"))

fourth_tail <- ggplot(data= filter(random_model, 
                               grepl("tail", rownames(random_model)) & 
                               Migration == 4), aes(Estimate)) + 
  geom_histogram(fill="black", bins = no.bin) +
  geom_vline(aes(xintercept = filter(clogit_results, variable == "tail" & 
                               Migration == 4)$coefsTCV), color="red", size=1) +
  theme_classic()+
  theme(legend.position = "none", text = element_text(size=20), 
        axis.text.x = element_text(colour = "black"),
        axis.text.y = element_text(colour = "black"))
fourth_cross <- ggplot(data= filter(random_model, 
                                grepl("cross", rownames(random_model)) 
                                & Migration == 4), aes(Estimate)) + 
  geom_histogram(fill="black", bins = no.bin) +
  geom_vline(aes(xintercept = filter(clogit_results, variable == "cross" & 
                               Migration == 4)$coefsTCV), color="red", size=1) +
  theme_classic()+
  theme(legend.position = "none", text = element_text(size=20), 
        axis.text.x = element_text(colour = "black"),
        axis.text.y = element_text(colour = "black"))
fourth_lift <- ggplot(data= filter(random_model, 
                                grepl("vertical", rownames(random_model)) 
                                & Migration == 4), aes(Estimate)) + 
  geom_histogram(fill="black", bins = no.bin) +
  geom_vline(aes(xintercept = filter(clogit_results, variable == "vertical" & 
                               Migration == 4)$coefsTCV), color="red", size=1) +
  theme_classic()+
  theme(legend.position = "none", text = element_text(size=20), 
        axis.text.x = element_text(colour = "black"),
        axis.text.y = element_text(colour = "black"))

adult_tail <- ggplot(data= filter(random_model, 
                               grepl("tail", rownames(random_model)) & 
                               Migration == "5 +"), aes(Estimate)) + 
  geom_histogram(fill="black", bins = no.bin) +
  geom_vline(aes(xintercept = filter(clogit_results, variable == "tail" & 
                               Migration == "5 +")$coefsTCV), color="red", size=1) +
  theme_classic()+
  theme(legend.position = "none", text = element_text(size=20), 
        axis.text.x = element_text(colour = "black"),
        axis.text.y = element_text(colour = "black"))
adult_cross <- ggplot(data= filter(random_model, 
                                grepl("cross", rownames(random_model)) 
                                & Migration == "5 +"), aes(Estimate)) + 
  geom_histogram(fill="black", bins = no.bin) +
  geom_vline(aes(xintercept = filter(clogit_results, variable == "cross" & 
                               Migration == "5 +")$coefsTCV), color="red", size=1) +
  theme_classic()+
  theme(legend.position = "none", text = element_text(size=20), 
        axis.text.x = element_text(colour = "black"),
        axis.text.y = element_text(colour = "black"))
adult_lift <- ggplot(data= filter(random_model, 
                                grepl("vertical", rownames(random_model)) 
                                & Migration == "5 +"), aes(Estimate)) + 
  geom_histogram(fill="black", bins = no.bin) +
  geom_vline(aes(xintercept = filter(clogit_results, variable == "vertical" & 
                               Migration == "5 +")$coefsTCV), color="red", size=1) +
  theme_classic()+
  theme(legend.position = "none", text = element_text(size=20), 
        axis.text.x = element_text(colour = "black"),
        axis.text.y = element_text(colour = "black"))

all_15 <- ggarrange(first_tail + rremove("xylab"), second_tail+ rremove("xylab"),
                    third_tail + rremove("xylab"), fourth_tail+ rremove("xylab"), 
                    adult_tail + rremove("xylab"), first_cross + rremove("xylab"),
                    second_cross + rremove("xylab"), third_cross+ rremove("xylab"),
                    fourth_cross + rremove("xylab"), adult_cross+ rremove("xylab"),
                    first_lift + rremove("ylab") + labs(x = "1"), second_lift +
                    rremove("ylab") + labs(x = "2"), third_lift+ rremove("ylab") + 
                    labs(x = "3"), fourth_lift+ rremove("ylab") + labs(x = "4"), 
                    adult_lift+ rremove("ylab") + labs(x = "5 +"),
                    ncol=5, nrow=3, legend = "none")

annotate_figure(all_15, left = text_grob("Frequency", rot = 90, vjust = 1,),
                bottom = text_grob("Coefficient"))
```
---
Comparisons of the observed coefficients (red) and the distribution of coefficients from randomized data (black) in each migratory journey (1 to 5 or more) and for each predictor.

